---
title: "Data manipulation"
output: 
  html_document:
    code_download: true
    toc: true
    toc_float: false
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

library(tidyverse)
penguins <- read_csv("data/penguins.csv")
```

## Creating summary tables

In spreadsheet pivot tables or dinamic tables are powerful tools for summarizing data in different ways. We can create these tables using the `group_by` and `summarize` functions from the Tidyverse. 

Let's take a look to our report, right after we read the data, we create a summary with the number of cases and the average body mass for each specie in our dataset.

We can use `group_by()` to group our data by **species** and `summarize()` to calculate the number of cases and the average **body_mass_g** for each grouping.

```{r}

penguins %>% 
  group_by(species) %>% 
  summarise(count = n(),
            mean_body_mass = mean(body_mass_g, na.rm = TRUE))

```
The function `n()` return the number of cases and the function `mean` calculate the average. The parameter `na.rm = TRUE` tells the function that remove the empty values, the `NA` values before made the calculation.


> Question: What if you don’t group_by first? Let’s try it and discuss what’s going on.

```{r}
penguins %>% 
  summarise(count = n(),
            mean_body_mass = mean(body_mass_g, na.rm = TRUE))
```

So if we don’t group_by first, we will get a single summary statistic (n and mean in this case) for the whole dataset.

Now, we want to have the average of the body mass in kilograms and with two decimals.  We can change our summarise to get that calculation.  

```{r}
penguins %>% 
  group_by(species) %>% 
  summarise(count = n(),
            mean_body_mass = round(mean(body_mass_g/1000, na.rm = TRUE), 2))
```

First we divide body mass by 1000 to transform from grams to kilograms and then we use the function `round` to get only two decimals.

Now we are at the point where we actually want to save this summary information as a variable so we can use it in further analyses and formatting.

So let’s add a variable assignment to that first line:

```{r}
summary <- penguins %>% 
  group_by(species) %>% 
  summarise(count = n(),
            mean_body_mass = round(mean(body_mass_g/1000, na.rm = TRUE), 2))
```

## Formatig the table

To make our table look better in a report, we have to change some of its elements. For example, the name of the variables should be informative for a person (at the moment they are useful names for a computer), we can add a title to it and change the alignment of the data.

We can do that with the `kable` function.

```{r}
summary %>% 
  knitr::kable(col.names = c("Specie", "Count", "Mean Body Mass"),
               align = "lcc",
               caption = "Summary of penguins on the data set by specie.")
```

We can do a lot more action with functions or verbs of the tidyverse.  Let's see a couple more

## Creating new variables

There are a lot of times where we want to operate beyond the original data. This is often done by adding a column. We do this with the `mutate()` function.

We can add to our counting example by using `mutate()` to create a new variable **prop**, which represents the proportion of penguins of each **species**

```{r}
penguins %>% 
  group_by(species) %>% 
  summarise(count = n(),
            mean_body_mass = round(mean(body_mass_g/1000, na.rm = TRUE), 2)) %>%
  mutate(prop = count/nrow(penguins))

```

## Filtering data

In our report we make plot by species. To do this we need to keep only the data of the specie we want to analyze.  We use the `filter` function to subset data frames, this functions return the rows that satisfy variable conditions.

We use expected operators (>, <, >=, <=, ==) to set conditions for a numeric variable when filtering.  This operwtor are also useful for date and character types of data.  Look at the code to see a condition for text type of data.

```{r}
penguins %>% 
  filter(species == "Gentoo")
```

> Activity: Write a sentence to filter all the penguins in the Torgersen island.

> Activity: Now write the code to filter all the penguins wich flipper lenght is greater than that 200