---
title: "Data manipulation"
output: 
  html_document:
    code_download: true
    toc: true
    toc_float: false
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

library(tidyverse)
penguins <- read_csv("data/penguins.csv")
```

## Creating summary tables

In spreadsheet pivot tables or dinamic tables are powerful tools for summarizing data in different ways. We can create these tables using the `group_by` and `summarize` functions from the Tidyverse. 

Let's take a look to our report, right after we read the data, we create a summary with the number of cases and the average body mass for each specie in our dataset.

We can use `group_by()` to group our data by **species** and `summarize()` to calculate the number of cases and the average **body_mass_g** for each grouping.

```{r}

penguins %>% 
  group_by(species) %>% 
  summarise(count = n(),
            mean_body_mass = mean(body_mass_g, na.rm = TRUE))

```
The function `n()` return the number of cases and the function `mean` calculate the average. The parameter `na.rm = TRUE` tells the function that remove the empty values, the `NA` values before made the calculation.


> Question: What if you don’t group_by first? Let’s try it and discuss what’s going on.

```{r}
penguins %>% 
  summarise(count = n(),
            mean_body_mass = mean(body_mass_g, na.rm = TRUE))
```

So if we don’t group_by first, we will get a single summary statistic (n and mean in this case) for the whole dataset.

Now, we want to have the average of the body mass in kilograms and with two decimals.  We can change our summarise to get that calculation.  

```{r}
penguins %>% 
  group_by(species) %>% 
  summarise(count = n(),
            mean_body_mass = round(mean(body_mass_g/1000, na.rm = TRUE), 2))
```

First we divide body mass by 1000 to transform from grams to kilograms and then we use the function `round` to get only two decimals.

Now we are at the point where we actually want to save this summary information as a variable so we can use it in further analyses and formatting.

So let’s add a variable assignment to that first line:

```{r}
summary <- penguins %>% 
  group_by(species) %>% 
  summarise(count = n(),
            mean_body_mass = round(mean(body_mass_g/1000, na.rm = TRUE), 2))
```

## Formatig the table

To make our table look better in a report, we have to change some of its elements. For example, the name of the variables should be informative for a person (at the moment they are useful names for a computer), we can add a title to it and change the alignment of the data.

We can do that with the `kable` function.

```{r}
summary %>% 
  knitr::kable(col.names = c("Specie", "Count", "Mean Body Mass"),
               align = "lcc",
               caption = "Summary of penguins on the data set by specie.")
```

