---
title: "Gentoo Penguins"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
library(readxl)
```

## About the data

The **Palmer Penguins** data were collected and made available by Dr. Kristen Gorman and the [Palmer Station, Antarctica LTER](https://pal.lternet.edu/), a member of the Long Term Ecological Research Network. The data set includes several characteristics from Adelie, Chinstrap and Gentoo penguins. 

```{r read}
penguins <- read_csv("data/penguins.csv")

summary <- penguins %>% 
  group_by(species) %>% 
  summarise(count = n(),
            mean_body_mass = round(mean(body_mass_g/1000, na.rm = TRUE), 2))

summary %>% 
  knitr::kable(col.names = c("Specie", "Count", "Mean Body Mass"),
               align = "lcc",
               caption = "Summary of penguins on the data set by species.")
```

### Gentoo penguins

On this section we focus the analysis on the Gentoo species. The bill length is positively correlated with bill depth, penguins with longer bills usually have deeper bills.

```{r}
penguins %>% 
  filter(species == "Gentoo") %>% 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(color = "darkorange",
             size = 3,
             alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "darkorange") +
  theme_minimal() +
  labs(title = "Penguin bill dimensions",
       subtitle = "Bill length and depth for Gentoo Penguins at Palmer Station LTER",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       color = "Penguin species") 
```

Female and male penguins have differnet body mass. Male penguins generally are bigger than females. 

```{r}
penguins %>% 
  filter(species == "Gentoo") %>% 
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(aes(fill = sex),
                 alpha = 0.5,
                 position = "identity") +
  scale_fill_manual(values = c("darkorange", "cyan4")) +
  theme_minimal() +
  labs(x = "Body mass (g)",
       y = "Frequency",
       fill = NULL,
       title = "Penguin body mass")
```

